---
title: "R Notebook"
output: html_notebook
---

BINF 6970 Assignment 2:
T cells are continuously generated in the bone marrow to maintain a diverse repertoire capable of responding to pathogens. Following thymic selection, naïve (NAI) T cells migrate to peripheral tissues, where they need to encounter specific antigens and additional signals for clonal expansion and differentiation into effector (EFFE) and memory (MEM) cells. The exact genetic mechanisms governing this differentiation process are not fully understood, and there is ongoing debate as to whether NAI T cells differentiate into EFFEs first, which subsequently become MEM cells (linear model), or if they directly differentiate into both EFFE and MEM cells simultaneously (parallel model). Consider the data provided in geneexpression2.rda where rows are samples from different subjects and different T cell types and the columns are a subset of gene expression measurements on the 156 most differentially expressed genes as studied in (Holmes et al., 2005).

Notes
- The row names (e.g. in HEA26_EFFE_1) provide important information:
- Hea26_ represents status of the subject and EFFE_1 represents a cell type.

1 - Perform PCA on this dataset treating differentially expressed genes as the variables.
2 - Visualize the PCs and investigate how they relate to cell and status types.
3 - Write down your interpretation considering the conclusions drawn in the article (Holmes et al., 2005)

```{r}
# Question 1 ----
# Load in file
load("geneexpression2.rda")
dat

# Load in libraries
library(tidyverse)
library(ggfortify)
library(dplyr)
library(patchwork)
library(ggrepel)

## Part 1 ----
## EDA
str(dat)
dim(dat) # 30 rows (patient samples) and 156 columns (differentially expressed genes)

# Check for summary statistics to determine if data is scaled
summary(dat)
colMeans(dat) # means are not 0 or close to 0
apply(dat, 2, sd) # standard deviations are not 1

# Check for NAs in file
sum(is.na(dat)) # 0 NA values in the dataset

# create column of type of cell/status for plotting
dat_name <- dat %>%
  rownames_to_column(var = "cell_type")

#Check if cell type column is added 
str(dat_name) 
dat_name$cell_type # cell type is present

#Grouping on cell stage and health status based on the information in the cell type data.
dat_name <- dat_name %>%
  mutate(status = case_when(str_starts(cell_type, "HEA") ~ "healthy", str_starts(cell_type, "MEL") ~ "melanoma"), type = case_when(str_detect(cell_type, "EFFE") ~ "effector", str_detect(cell_type, "MEM") ~ "memory", str_detect(cell_type, "NAI") ~ "naive"))

table(dat_name$status) #15 healthy, 15 melanonma, total of 30
table(dat_name$type) # 10 effector, 9 memory, 11 naive for a total of 30 samples

dat_name %>%
  group_by(cell_type) %>%
  summarise()

# Running PCA
pc_data_scaled <- prcomp(dat, scale. = T) ## prcomp has center = TRUE as default, which we will leave as the default

# Initial visualization of PCA results - distribution of how much variance is captured by each PC
plot(pc_data_scaled)

# Checks the variance explained by each PC (top 10)
round(head(summary(pc_data_scaled)[["importance"]]["Proportion of Variance",], 10), 4)

## Part 2 ----
# Scaled PCA plot with cell type and health status 
pca_plot_type_status <- autoplot(prcomp(dat, scale. = T), data = dat_name, colour = "type", shape = "status", size = 3) +
  theme_minimal() +
  labs(colour = "Cell Type", shape = "Status", title = "C")

pca_plot_type <- autoplot(prcomp(dat, scale. = T), data = dat_name, colour = "type", size = 3) +
  theme_minimal() + 
  labs(colour = "Cell Type", title = "A")

pca_plot_status <- autoplot(prcomp(dat, scale. = T), data = dat_name, shape = "status", size = 3) +
  theme_minimal() + 
  labs(shape = "Status", title = "B")

pca_plot_combined <- (pca_plot_type | pca_plot_status) / pca_plot_type_status
pca_plot_combined

## Screeplot
plot(as.ts(pc_data_scaled$sdev^2),
  ylab = "Variance", 
  xlab = "Principal Component") + 
axis(side = 1, at = seq(1, 30, by = 1), labels = FALSE) # adds ticks for x-axis

## Part 3 ----
# Extract loadings
# PC1
loadings <- pc_data_scaled$rotation
pc1_genes <- round(sort(abs(loadings[,1]), decreasing = TRUE), 4)
head(pc1_genes, 10)

#PC2 
pc2_genes <- round(sort(abs(loadings[,2]), decreasing = TRUE), 4)
head(pc2_genes, 10)

## Visualize all loadings and their contributions to variation in PC1 and PC2 with biplot
pca_biplot <- autoplot(pc_data_scaled, data = dat_name,
         loadings = TRUE,
         loadings.label = FALSE, 
         size = -1) +
  geom_text_repel(aes(x = PC1, y = PC2, 
                      label = case_when(
                        str_detect(cell_type, "EFFE") ~ "effector",
                        str_detect(cell_type, "MEM")  ~ "memory",
                        str_detect(cell_type, "NAI")  ~ "naive")), 
                  size = 3) +
  theme_minimal()

```

Cytokines storm, a hyperinflammatory state caused by a deregulated immune system, is a key
aspect of the novel coronavirus disease, with affected patients showing high levels of pro-
inflammatory cytokines, which correlate with disease severity (Vanderkamp et al., 2025).
Additional details regarding the role of cytokines in mediating the immune response can be
found Table 1 (see Appendix). For this problem, you are provided with a dataset that includes
measurements of 28 distinct analytes (cytokines, chemokines, and acute phase proteins) that
were obtained from plasma samples of 76 human patients who tested positive for COVID-19.


1- Develop a regularized logistic-regression model via Elastic-Net framework to predict
COVID-19 disease progression by analyzing cytokine profiles and patient characteristics.

1. Which cytokines are most effective at predicting the COVID-19 disease progression in
patients?

ie which are the top predictors?

2. Which cytokines show a significant association with COVID-19 disease severity?
3. How does patient age correlate with COVID-19 disease severity?

```{r}
library(glmnet)     # For regularization methods
library(caret)      # For data splitting and preprocessing
library(ggplot2)    # For visualization
library(dplyr)      # For data manipulation
library(pROC)       # For ROC calculations
library(patchwork)  # For combining plots
```

2- For model evaluation, randomly split the full dataset into a 75% training set and a 25% test
set (use seed: 1717 for reproducibility).

3- For model training, use seq (0, 1, .01) as the search grid for alpha, i.e. Ridge and
Lasso are included in the set of candidate elastic-net models. You may choose to use either
“auc” or “deviance” as the accuracy measure (loss function) in cv.glmnet at the training
stage (only use one of these two measure).

4- Perform two separate sets of analysis by considering 10-fold and 20-fold cross-validation.

5- Discuss the implications of cross-validation schemes (10/20 folds) on your analyses by
providing a clear and concise comparison of model selection and predictive performance.

Full workflow:

```{r}
# Setting the seed
set.seed(1717)

covid_data <- Immunologic_profiles_of_patients_with_COVID_19

# Load data
covid_data <- read.csv("covid.csv")

#Check fo NAs
sum(is.na(covid_data)) # is 0

# Convert y to a binary variable (factor class) where 0= mild and 1= severe to 
covid_data <- covid_data %>%
  mutate(severity_binary = as.factor(case_when(Severirty == "Mild" ~ 0, Severirty == "Severe" ~ 1)))

#Check for class type
class(covid_data$severity_binary)

# Separate into x and y
X <- as.matrix(covid_data[c(2, 5:32)])  # Predictors (expanded features)= everything but patient number and severity obviously # also omit sex
y <- covid_data$severity_binary # Response= severity
class(y)

# Setting the seed
set.seed(1717)

# Creating the test and training set, X is all the predictors and y is the response
train_index <- createDataPartition(y, p = 0.75, list = FALSE) # how to split randomly?- i think its built in
X_train <- X[train_index, ]
X_test <- X[-train_index, ]
y_train <- y[train_index]
y_test <- y[-train_index]

#Check for class 
class(y_train)
str(y_train)
```

```{r}
plot(as.ts(apply(X,2,sd)), xlab="covariate", ylab="sd")
# predictors need to be scaled 
```

```{r}
# Standardize predictors (glmnet does this by default with standardize=TRUE)
# But we'll scale manually for consistency across methods
X_train_scaled <- scale(X_train)
X_test_scaled <- scale(X_test, 
                       center = attr(X_train_scaled, "scaled:center"),
                       scale = attr(X_train_scaled, "scaled:scale"))

#Check the plot again after scaling
plot(as.ts(apply(X_train_scaled,2,sd)), xlab="covariate", ylab="sd") #SD is 1
plot(as.ts(apply(X_test_scaled,2,sd)), xlab="covariate", ylab="sd") # SD flunctuates but is mostly between 0 - 2
```

```{r}
# Now do the grid search for the best alpha
alpha_grid <- seq(0, 1, by = 0.01) # as specified

# Cross-validation
## using k=10 folds
# Create container for results
results_k10 <- data.frame()

# Perform grid search, this will test all the values of alpha then we'll be able to see which one was best
for (alpha_val in alpha_grid) {
  # Cross-validation for current alpha
  cv_model_k10 <- cv.glmnet(X_train_scaled, y_train, 
                        alpha = alpha_val, 
                        nfolds = 10,
                        type.measure = "auc", # using AUC for logistic
                        keep = TRUE,
                        family="binomial")
  
  # Store results
  results_k10 <- rbind(results_k10, data.frame(
    alpha = alpha_val,
    lambda_min = cv_model_k10$lambda.min,
    lambda_1se = cv_model_k10$lambda.1se,
    cvm_min = min(cv_model_k10$cvm),
    cvm_1se = cv_model_k10$cvm[which(cv_model_k10$lambda == cv_model_k10$lambda.1se)],
    nzero_min = cv_model_k10$nzero[which(cv_model_k10$lambda == cv_model_k10$lambda.min)],
    nzero_1se = cv_model_k10$nzero[which(cv_model_k10$lambda == cv_model_k10$lambda.1se)]
  ))
}


# error Warning: Too few (< 10) observations per fold for type.measure='auc' in cv.lognet; changed to type.measure='deviance'. Alternatively, use smaller value for nfolds
```

```{r}
# Plot
plot(cv_model_k10)
```

```{r}
# Print results
print(results_k10)

# Find best alpha based on minimum CV error
best_alpha_min_k10 <- results_k10$alpha[which.min(results_k10$cvm_min)] # 0.42
best_alpha_1se_k10 <- results_k10$alpha[which.min(results_k10$cvm_1se)] # 0.51

# extract lambda
lambda_min_k10 <- results_k10$lambda_min[results_k10$alpha == best_alpha_min_k10] #0.0485
lambda_1se_k10 <- results_k10$lambda_1se[results_k10$alpha == best_alpha_1se_k10] #0.177

# Fit final models for both min and 1se
enet_model_min_k10 <- glmnet(X_train_scaled, y_train, 
                         alpha = best_alpha_min_k10, 
                         lambda = lambda_min_k10,
                         family = "binomial",
                         keep = "TRUE")

enet_model_1se_k10 <- glmnet(X_train_scaled, y_train, 
                         alpha = best_alpha_1se_k10, 
                         lambda = lambda_1se_k10,
                         family = "binomial",
                         keep = "TRUE")

# Fitted values and test set predictions for lambda.min
prds.train_min_k10 <- predict(enet_model_min_k10,newx = X_train_scaled, type = "response", s=lambda_min_k10)[,1]
prds.test_min_k10 <- predict(enet_model_min_k10,newx = X_test_scaled, type = "response", s=lambda_min_k10)[,1]

# Prediction Accuracy
auc.train_min_k10 <- roc(y_train,prds.train_k10)
auc.train_min_k10 #0.9634

auc.test_min_k10 <- roc(y_test,prds.test_k10)
auc.test_min_k10 #0.8409

# Fitted values and test set predictions for k10, lambda.1se
prds.train_1se_k10 <- predict(enet_model_1se_k10,newx = X_train_scaled, type = "response", s=lambda_1se_k10)[,1]
prds.test_1se_k10 <- predict(enet_model_1se_k10,newx = X_test_scaled, type = "response", s=lambda_1se_k10)[,1]

# Prediction Accuracy
auc.train_1se_k10 <- roc(y_train,prds.train_1se)
auc.train_1se_k10 #0.9078

auc.test_1se_k10 <- roc(y_test,prds.test_1se)
auc.test_1se_k10 #0.7614
```

```{r}
#plot AUC curves to visualize for k fold = 10, lambda.min
par(mfrow=c(1,2))
plot(auc.train_min_k10)
plot(auc.test_min_k10)

#for k fold 10, lambda.1se
par(mfrow=c(1,2))
plot(auc.train_1se_k10)
plot(auc.test_1se_k10) 
```

```{r}
# Confusion matrix for k10 min
table(y_train)
table(y_test) 
table(as.numeric(prds.train_min_k10>.5))

conf.mat_min_k10 <- table(y=y_train,yhat=as.numeric(prds.train_min_k10>.5))
conf.mat_min_k10

## A function to compute Sensitivity, Specificity, and accuracy
sn.sp.acc <- function(mat){
  sn <- round(mat[2,2]/sum(mat[2,]),4)
  sp <- round(mat[1,1]/sum(mat[1,]),4)
  acc <- round((mat[1,1] + mat[2,2]) / sum(mat), 4)
  return(unlist(list(sensitivity=sn, specificity=sp, accuracy=acc)))
}

sn.sp.acc(conf.mat_min_k10)
# this was for threshold 0.5
#sensitivity 0.939
#specificity 0.917
#accuracy 0.930

# Need to find threshold to balance between sensitivity and specificity
# function to find threshold
threshold <- function (auc_input) {
  snsp.train <- cbind(auc_input$sensitivities,auc_input$specificities)
  head(snsp.train)

  indx <- which.max(apply(snsp.train,1,min))  ### min-max approach

  ## the threshold
  cutoff <- auc_input$thresholds[indx]
  return(cutoff)
}

# for training
cutoff_train_min_k10 <- round(threshold(auc.train_min_k10),4)
cutoff_train_min_k10 #0.4868

#Confusion matrix for new cutoff value
table(y=y_train,yhat=as.numeric(prds.train_min_k10>cutoff))
sn.sp.acc(table(y=y_train,yhat=as.numeric(prds.train_min_k10>cutoff))) ## TRAIN SET
#sensitivity specificity accuracy 
#     0.9697       0.9167     0.9474
      
# For testing set
cutoff_test_min_k10 <- round(threshold(auc.test_min_k10),3)
cutoff_test_min_k10 #0.5840

#Confusion matrix table for new cutoff
table(y=y_test,yhat=as.numeric(prds.test_min_k10>cutoff_test))
sn.sp.acc(table(y=y_test,yhat=as.numeric(prds.test_min_k10>cutoff_test))) ## TEST SET
#sensitivity specificity  accuracy
#      0.8182       0.7500     0.7895
```

```{r}
# Confusion matrix for k10 lambda.1se
# Need to balance between sensitivity and specificity
# for training
cutoff_train_1se_k10 <- round(threshold(auc.train_1se_k10),4)
cutoff_train_1se_k10 #0.6059

#Confusion matrix for new cutoff value
table(y=y_train,yhat=as.numeric(prds.train_1se_k10>cutoff_train_1se_k10))
sn.sp.acc(table(y=y_train,yhat=as.numeric(prds.train_1se_k10>cutoff_train_1se_k10))) ## TRAIN SET
# sensitivity specificity    accuracy 
#      0.8788       0.9167       0.8947 
      
# For testing set
## finding the threshold
cutoff_test_1se_k10 <- round(threshold(auc.test_1se_k10),4)
cutoff_test_1se_k10 #0.5993

#Confusion matrix table for new cutoff
table(y=y_test,yhat=as.numeric(prds.test_1se_k10>cutoff_test_1se))
sn.sp.acc(table(y=y_test,yhat=as.numeric(prds.test_1se_k10>cutoff_test_1se))) ## TEST SET
#sensitivity specificity    accuracy 
#      0.6364       0.7500       0.6842 

```

```{r}
## with k=20 ###
results_k20 <- data.frame()

# Perform grid search, this will test all the values of alpha then we'll be able to see which one was best
for (alpha_val in alpha_grid) {
  # Cross-validation for current alpha
  cv_model_k20 <- cv.glmnet(X_train_scaled, y_train, 
                        alpha = alpha_val, 
                        nfolds = 20,
                        type.measure = "auc", # either auc or deviance, not sure the difference
                        keep = TRUE,
                        family = "binomial")
  
  # Store results
  results_k20 <- rbind(results_k20, data.frame(
    alpha = alpha_val,
    lambda_min = cv_model_k20$lambda.min,
    lambda_1se = cv_model_k20$lambda.1se,
    cvm_min = min(cv_model_k20$cvm),
    cvm_1se = cv_model_k20$cvm[which(cv_model_k20$lambda == cv_model_k20$lambda.1se)],
    nzero_min = cv_model_k20$nzero[which(cv_model_k20$lambda == cv_model_k20$lambda.min)],
    nzero_1se = cv_model_k20$nzero[which(cv_model_k20$lambda == cv_model_k20$lambda.1se)]
  ))
}
```

```{r}
# Plot
plot(cv_model_k20)
```
```{r}
# Print results
print(results_k20)

# Find best alpha based on minimum CV error
best_alpha_min_k20 <- results_k20$alpha[which.min(results_k20$cvm_min)] # 0.08
best_alpha_1se_k20 <- results_k20$alpha[which.min(results_k20$cvm_1se)] # 0.42

#Extract best lambda 
lambda_min_k20 <- results_k20$lambda_min[results_k20$alpha == best_alpha_min_k20] #0.0477
lambda_1se_k20 <- results_k20$lambda_1se[results_k20$alpha == best_alpha_1se_k20] #0.2110

# Fit best model using glmnet
# Fit final models kfold 20
enet_model_min_k20 <- glmnet(X_train_scaled, y_train, 
                         alpha = best_alpha_min_k20, 
                         lambda = lambda_min_k20,
                         family = "binomial",
                         keep = "TRUE")

enet_model_1se_k20 <- glmnet(X_train_scaled, y_train, 
                         alpha = best_alpha_1se_k20, 
                         lambda = lambda_1se_k20,
                         family = "binomial",
                         keep = "TRUE")

# Fitted values and test set predictions for k10 lambda.min
prds.train_min_k20 <- predict(enet_model_min_k20,newx = X_train_scaled, type = "response", s=lambda_min_k20)[,1]
prds.test_min_k20 <- predict(enet_model_min_k20,newx = X_test_scaled, type = "response", s=lambda_min_k20)[,1]

prds.train_1se_k20 <- predict(enet_model_1se_k20,newx = X_train_scaled, type = "response", s=lambda_1se_k20)[,1]
prds.test_1se_k20 <- predict(enet_model_1se_k20,newx = X_test_scaled, type = "response", s=lambda_1se_k20)[,1]

# Prediction Accuracy
auc.train_min_k20 <- roc(y_train,prds.train_min_k20)
auc.train_min_k20 #0.9596

auc.test_min_k20 <- roc(y_test,prds.test_min_k20)
auc.test_min_k20 #0.8068

auc.train_1se_k20 <- roc(y_train, prds.train_1se_k20)
auc.train_1se_k20 #0.9053

auc.test_1se_k20 <- roc(y_test, prds.test_1se_k20)
auc.test_1se_k20 #0.7614
```

```{r}
par(mfrow=c(1,2))
plot(auc.train_min_k20)
plot(auc.test_min_k20)

par(mfrow=c(1,2))
plot(auc.train_1se_k20)
plot(auc.test_1se_k20)
```

```{r}
# Confusion matrix
table(y_train)
table(y_test) 
table(as.numeric(prds.train_min_k20>.5))

conf.mat_min_k20 <- table(y=y_train,yhat=as.numeric(prds.train_min_k20>.5))
conf.mat_min_k20

sn.sp.acc <- function(mat){
  sn <- round(mat[2,2]/sum(mat[2,]),4)
  sp <- round(mat[1,1]/sum(mat[1,]),4)
  acc <- round((mat[1,1] + mat[2,2]) / sum(mat), 4)
  return(unlist(list(sensitivity=sn, specificity=sp, accuracy=acc)))
}

sn.sp.acc(conf.mat_min_k20)
#sensitivity 0.970
#specificity 0.875
#accuracy 0.930

```

```{r}
# Confusion matrix for k20 lambda min
# Need to balance between sensitivity and specificity
# for training
cutoff_train_min_k20 <- round(threshold(auc.train_min_k20),4)
cutoff_train_min_k20 #0.5438

#Confusion matrix for new cutoff value
table(y=y_train,yhat=as.numeric(prds.train_min_k20>cutoff_train_min_k20))
sn.sp.acc(table(y=y_train,yhat=as.numeric(prds.train_min_k20>cutoff_train_min_k20))) ## TRAIN SET
# sensitivity specificity    accuracy 
#      0.9394       0.9167        0.9298
      
# For testing set
## finding the threshold
cutoff_test_min_k20 <- round(threshold(auc.test_min_k20),4)
cutoff_test_min_k20 #0.5608

#Confusion matrix table for new cutoff
table(y=y_test,yhat=as.numeric(prds.test_min_k20>cutoff_test_min_k20))
sn.sp.acc(table(y=y_test,yhat=as.numeric(prds.test_min_k20>cutoff_test_min_k20))) ## TEST SET
#sensitivity specificity    accuracy 
#      0.8182       0.7500       0.7895
```

```{r}
# Confusion matrix for k20 lambda 1se
# Need to balance between sensitivity and specificity
# for training
cutoff_train_1se_k20 <- round(threshold(auc.train_1se_k20),4)
cutoff_train_1se_k20 #0.6040

#Confusion matrix for new cutoff value
table(y=y_train,yhat=as.numeric(prds.train_1se_k20>cutoff_train_1se_k20))
sn.sp.acc(table(y=y_train,yhat=as.numeric(prds.train_1se_k20>cutoff_train_1se_k20))) ## TRAIN SET
#sensitivity specificity    accuracy 
#      0.8788       0.9167       0.8947 
      
# For testing set
## finding the threshold
cutoff_test_1se_k20 <- round(threshold(auc.test_1se_k20),4)
cutoff_test_1se_k20 #0.5957

#Confusion matrix table for new cutoff
table(y=y_test,yhat=as.numeric(prds.test_1se_k20>cutoff_test_1se_k20))
sn.sp.acc(table(y=y_test,yhat=as.numeric(prds.test_1se_k20>cutoff_test_1se_k20))) ## TEST SET
#sensitivity specificity    accuracy 
#      0.6364       0.7500       0.6842
```

```{r}
# Combine ROC plots to compare 10- and 20- fold cross validation for testing data with lambda min
p1 <- ggroc(auc.test_min_k10, col = "blue") + 
  ggtitle(paste0("ROC Curve (10-fold CV)  AUC = ", round(auc(auc.test_min_k10),4)))

p2 <- ggroc(auc.test_min_k20, col = "red") + 
  ggtitle(paste0("ROC Curve (20-fold CV)  AUC = ", round(auc(auc.test_min_k20),4)))

p1 + p2
```
```{r}
# Combine ROC plots to compare 10- and 20- fold cross validation for testing data with lambda 1se
p3 <- ggroc(auc.test_1se_k10, col = "blue") + 
  ggtitle(paste0("ROC Curve (10-fold CV)  AUC = ", round(auc(auc.test_1se_k10),4)))

p4 <- ggroc(auc.test_1se_k20, col = "red") + 
  ggtitle(paste0("ROC Curve (20-fold CV)  AUC = ", round(auc(auc.test_1se_k20),4)))

p3 + p4
```


```{r}
# Combine ROC plots to compare 10- and 20- fold cross validation for training data with lambda min
p5 <- ggroc(auc.train_min_k10, col = "blue") + 
  ggtitle(paste0("ROC Curve (10-fold CV)  AUC = ", round(auc(auc.train_min_k10),4)))

p6 <- ggroc(auc.train_min_k20, col = "red") + 
  ggtitle(paste0("ROC Curve (20-fold CV)  AUC = ", round(auc(auc.train_min_k20),4)))

p5 + p6
```
```{r}
# Combine ROC plots to compare 10- and 20- fold cross validation for training data with lambda min
p7 <- ggroc(auc.train_1se_k10, col = "blue") + 
  ggtitle(paste0("ROC Curve (10-fold CV)  AUC = ", round(auc(auc.train_1se_k10),4)))

p8 <- ggroc(auc.train_1se_k20, col = "red") + 
  ggtitle(paste0("ROC Curve (20-fold CV)  AUC = ", round(auc(auc.train_1se_k20),4)))

p7 + p8
```


```{r}
## extract top predictors
# Function to extract top N predictors from a model
extract_top_predictors <- function(model, n = 10, lambda_type = type) {
  # Get coefficients
  if (lambda_type == "min") {
    coefs <- as.matrix(coef(model))[, 1]
  } else {
    # For models stored differently
    coefs <- as.matrix(coef(model, s = paste0("lambda.", lambda_type)))[, 1]
  }
  
  # Remove intercept
  coefs_no_intercept <- coefs[-1]
  
  # Get absolute values for ranking
  abs_coefs <- abs(coefs_no_intercept)
  
  # Get top N predictors
  top_indices <- order(abs_coefs, decreasing = TRUE)[1:min(n, length(abs_coefs))]
  
  # Create result data frame
  result <- data.frame(
    Predictor = names(coefs_no_intercept)[top_indices],
    Coefficient = coefs_no_intercept[top_indices],
    Absolute_Effect = abs_coefs[top_indices],
    Rank = 1:length(top_indices)
  )
  
  return(result)
}

# Extract top predictors for each model
cat("\n--- Top 10 Predictors (Lambda.min) ---\n")

#For k fold 10
cat("\nElastic Net (lambda.min):\n")
enet_top_min_k10 <- extract_top_predictors(enet_model_min_k10, n = 10, lambda_type = "min")
print(enet_top_min_k10)

cat("\nElastic Net (lambda.1se):\n")
enet_top_1se_k10 <- extract_top_predictors(enet_model_1se_k10, n = 10, lambda_type = "1se")
print(enet_top_1se_k10)

#For kfold 20
enet_top_min_k20 <- extract_top_predictors(enet_model_min_k20, n = 10, lambda_type = "min")
print(enet_top_min_k20)

enet_top_1se_k20 <- extract_top_predictors(enet_model_1se_k20, n = 10, lambda_type = "1se")
print(enet_top_1se_k20)

```

To see which are the top predictors:
look at section 5 in regularisation methods comparison.R
