---
title: "R Notebook"
output: html_notebook
---

BINF 6970 Assignment 2:

1- Perform PCA on this dataset treating differentially expressed genes as the variables.

```{r}
# Load in file
load("geneexpression2.rda")
dat

# Load in libraries
library(tidyverse)
library(ggfortify)
library(dplyr)

## EDA
str(dat)
dim(dat) # 30 rows (patient samples) and 156 columns (differentially expressed genes)

# Check for NAs in file
sum(is.na(dat)) # 0 NA values in the dataset

# create column of type of cell/status for plotting
dat_name <- dat %>%
  rownames_to_column(var = "cell_type")

#Check if cell type column is added 
str(dat_name) 
dat_name$cell_type # cell type is present

dat_name <- dat %>%
  rownames_to_column(var = "cell_type") %>%
  mutate(status = case_when(cell_type(starts_with("HEA"))))

#Grouping on cell stage and health status based on the information in the cell type data.
dat_name <- dat_name %>%
  mutate(status = case_when(str_starts(cell_type, "HEA") ~ "healthy", str_starts(cell_type, "MEL") ~ "melanoma"), type = case_when(str_detect(cell_type, "EFFE") ~ "effector", str_detect(cell_type, "MEM") ~ "memory", str_detect(cell_type, "NAI") ~ "naive"))

table(dat_name$status) #15 healthy, 15 melanonma, total of 30
table(dat_name$type) # 10 effector, 9 memory, 11 naive for a total of 30 samples

dat_name %>%
  group_by(cell_type) %>%
  summarise()

# Running PCA

pc_data <- prcomp(dat) #, cor = F, scores = T) ## should we scale data?

pc_data_scaled <- prcomp(dat, scale. = T)

plot(pc_data)

#Checks the variance explained by each PC
summary(pc_data)[["importance"]]["Proportion of Variance",]

## Plotting scree plot and PC plot
# plot all 
autoplot(prcomp(dat), data = dat_name, colour="type", label = F, label.size = 3, size = 3, shape = "status") + theme_minimal()+labs(colour = "Cell Type", shape = "Status")  

## screeplot
plot(as.ts(pc_data$sdev^2)) +
axis(side = 1, at = seq(1, 30, by = 1), labels = FALSE) # adds ticks for x-axis

autoplot(prcomp(dat), data = dat_name, colour="status", label = F, label.size = 3, point.size= 5) + theme_minimal()+scale_colour_discrete(name="Status")

## No scaling because every predictor is in same unit + DE data should be already scaled 
## Plot scaled 
autoplot(prcomp(dat, scale. = T), data = dat_name, colour="type", label = F, label.size = 3, size = 3, shape = "status") + theme_minimal()+labs(colour = "Cell Type", shape = "Status")  

## need to separate by T cell type and status- not sure exactly what this is

# Status= HEA -healthy and MEL=melanoma

# separate columns






```


2- Visualize the PCs and investigate how they relate to cell and status types.
3- Write down your interpretation considering the conclusions drawn in the article (Holmes et
al., 2005)

```{r}

```



Cytokines storm, a hyperinflammatory state caused by a deregulated immune system, is a key
aspect of the novel coronavirus disease, with affected patients showing high levels of pro-
inflammatory cytokines, which correlate with disease severity (Vanderkamp et al., 2025).
Additional details regarding the role of cytokines in mediating the immune response can be
found Table 1 (see Appendix). For this problem, you are provided with a dataset that includes
measurements of 28 distinct analytes (cytokines, chemokines, and acute phase proteins) that
were obtained from plasma samples of 76 human patients who tested positive for COVID-19.


1- Develop a regularized logistic-regression model via Elastic-Net framework to predict
COVID-19 disease progression by analyzing cytokine profiles and patient characteristics.

1. Which cytokines are most effective at predicting the COVID-19 disease progression in
patients?

ie which are the top predictors?

2. Which cytokines show a significant association with COVID-19 disease severity?
3. How does patient age correlate with COVID-19 disease severity?

```{r}
library(lars)       # For diabetes dataset
library(glmnet)     # For regularization methods
library(caret)      # For data splitting and preprocessing
library(ggplot2)    # For visualization
library(dplyr)      # For data manipulation

# Set seed for reproducibility
set.seed(1717)

# Load data


# Check dimensions
cat("Dataset dimensions:", dim(X), "\n")
cat("Number of predictors:", ncol(X), "\n")
cat("Number of observations:", length(y), "\n")

# Split data into training (75%) and testing (25%) # how to randomly split?
train_index <- createDataPartition(y, p = 0.75, list = FALSE)
X_train <- X[train_index, ]
X_test <- X[-train_index, ]
y_train <- y[train_index]
y_test <- y[-train_index]

# Standardize predictors (glmnet does this by default with standardize=TRUE)
# But we'll scale manually for consistency across methods
X_train_scaled <- scale(X_train)
X_test_scaled <- scale(X_test, 
                       center = attr(X_train_scaled, "scaled:center"),
                       scale = attr(X_train_scaled, "scaled:scale"))

# Remove any columns with NA values (if scaling creates them due to zero variance)
na_cols <- which(colSums(is.na(X_train_scaled)) > 0)
if (length(na_cols) > 0) {
  X_train_scaled <- X_train_scaled[, -na_cols]
  X_test_scaled <- X_test_scaled[, -na_cols]
  cat("Removed", length(na_cols), "columns with zero variance\n")
}

# now do the elasticnet

# Cross-validation for elastic net with alpha=0.5
cv_enet <- cv.glmnet(X_train_scaled, y_train, 
                     alpha = 0.5, 
                     nfolds = 10,
                     type.measure = "mse")

plot(cv_enet, main = "Elastic Net (alpha=0.5) - CV Error")
# how many predictor left?
# Extract lambda values
enet_lambda_min <- cv_enet$lambda.min
enet_lambda_1se <- cv_enet$lambda.1se

cat("Elastic Net - Lambda min:", enet_lambda_min, "\n")
cat("Elastic Net - Lambda 1se:", enet_lambda_1se, "\n")

# Fit final models
enet_model_min <- glmnet(X_train_scaled, y_train, 
                         alpha = 0.5, 
                         lambda = enet_lambda_min)

enet_model_1se <- glmnet(X_train_scaled, y_train, 
                         alpha = 0.5, 
                         lambda = enet_lambda_1se)


```



2- For model evaluation, randomly split the full dataset into a 75% training set and a 25% test
set (use seed: 1717 for reproducibility).


3- For model training, use seq (0, 1, .01) as the search grid for alpha, i.e. Ridge and
Lasso are included in the set of candidate elastic-net models. You may choose to use either
“auc” or “deviance” as the accuracy measure (loss function) in cv.glmnet at the training
stage (only use one of these two measure).

grid search:

```{r}
alpha_grid <- seq(0, 1, by = 0.01)

# Create container for results
results <- data.frame()

# Perform grid search
for (alpha_val in alpha_grid) {
  # Cross-validation for current alpha
  cv_model <- cv.glmnet(X_train_scaled, y_train, 
                        alpha = alpha_val, 
                        nfolds = 10,
                        type.measure = "mse",
                        keep = TRUE)
  
  # Store results
  results <- rbind(results, data.frame(
    alpha = alpha_val,
    lambda_min = cv_model$lambda.min,
    lambda_1se = cv_model$lambda.1se,
    cvm_min = min(cv_model$cvm),
    cvm_1se = cv_model$cvm[which(cv_model$lambda == cv_model$lambda.1se)],
    nzero_min = cv_model$nzero[which(cv_model$lambda == cv_model$lambda.min)],
    nzero_1se = cv_model$nzero[which(cv_model$lambda == cv_model$lambda.1se)]
  ))
}

# Print results
print(results)

# Find best alpha based on minimum CV error
best_alpha_min <- results$alpha[which.min(results$cvm_min)]
best_alpha_1se <- results$alpha[which.min(results$cvm_1se)]

cat("\nBest alpha (lambda.min):", best_alpha_min, "\n")
cat("Best alpha (lambda.1se):", best_alpha_1se, "\n")

# Visualize alpha grid search
ggplot(results, aes(x = alpha)) +
  geom_line(aes(y = cvm_min, color = "lambda.min"), size = 1.2) +
  geom_line(aes(y = cvm_1se, color = "lambda.1se"), size = 1.2) +
  geom_point(aes(y = cvm_min, color = "lambda.min"), size = 3) +
  geom_point(aes(y = cvm_1se, color = "lambda.1se"), size = 3) +
  labs(title = "CV Error vs Alpha",
       x = "Alpha (0=Ridge, 1=Lasso)",
       y = "CV Error (MSE)",
       color = "Lambda Type") +
  theme_minimal()
```


4- Perform two separate sets of analysis by considering 10-fold and 20-fold cross-validation.
cv_enet <- cv.glmnet(X_train_scaled, y_train, 
                     alpha = 0.5, 
                     nfolds = 10,
                     type.measure = "mse")

cv_enet <- cv.glmnet(X_train_scaled, y_train, 
                     alpha = 0.5, 
                     nfolds = 20,
                     type.measure = "mse")


5- Discuss the implications of cross-validation schemes (10/20 folds) on your analyses by
providing a clear and concise comparison of model selection and predictive performance.

```{r}

```




Full workflow:

```{r}
# Setting the seed
set.seed(1717)

# Load data

covid_data <- read.csv("covid.csv")

sum(is.na(covid_data))

# Convert y to a binary variable where 0= mild and 1= severe
covid_data <- covid_data %>%
  mutate(severity_binary = as.factor(case_when(Severirty == "Mild" ~ 0, Severirty == "Severe" ~ 1)))

class(covid_data$severity_binary)

# Separate into x and y

X <- as.matrix(covid_data[c(2, 5:32)])  # Predictors (expanded features)= everything but patient number and severity obviously # also omit sex
y <- covid_data$severity_binary # Response= severity
class(y)


# Creating the test and training set, X is all the predictors and y is the response

train_index <- createDataPartition(y, p = 0.75, list = FALSE) # how to split randomly?- i think its built in
X_train <- X[train_index, ]
X_test <- X[-train_index, ]
y_train <- y[train_index]
y_test <- y[-train_index]

class(y_train)
str(y_train)
# Now do the grid search for the best alpha

## first we do it with k=10 folds

alpha_grid <- seq(0, 1, by = 0.01) # as specified

# Create container for results
results_k10 <- data.frame()

# manually standardise predictors

# Standardize predictors (glmnet does this by default with standardize=TRUE)
# But we'll scale manually for consistency across methods
X_train_scaled <- scale(X_train)
X_test_scaled <- scale(X_test, 
                       center = attr(X_train_scaled, "scaled:center"),
                       scale = attr(X_train_scaled, "scaled:scale"))



# Perform grid search, this will test all teh values of alpha then we'll be able to see which one was best
for (alpha_val in alpha_grid) {
  # Cross-validation for current alpha
  cv_model <- cv.glmnet(X_train_scaled, y_train, 
                        alpha = alpha_val, 
                        nfolds = 10,
                        type.measure = "auc", # using AUC for logistic
                        keep = TRUE,
                        family="binomial")
                         # instead of manually
  
  # Store results
  results_k10 <- rbind(results_k10, data.frame(
    alpha = alpha_val,
    lambda_min = cv_model$lambda.min,
    lambda_1se = cv_model$lambda.1se,
    cvm_min = min(cv_model$cvm),
    cvm_1se = cv_model$cvm[which(cv_model$lambda == cv_model$lambda.1se)],
    nzero_min = cv_model$nzero[which(cv_model$lambda == cv_model$lambda.min)],
    nzero_1se = cv_model$nzero[which(cv_model$lambda == cv_model$lambda.1se)]
  ))
}

# error Warning: Too few (< 10) observations per fold for type.measure='auc' in cv.lognet; changed to type.measure='deviance'. Alternatively, use smaller value for nfolds

# Print results
print(results_k10)

# Find best alpha based on minimum CV error
best_alpha_min <- results_k10$alpha[which.min(results_k10$cvm_min)] # 0.72
best_alpha_1se <- results_k10$alpha[which.min(results_k10$cvm_1se)] # 0.69

# extract lambda when alpha=0.72
lambda_min <- results_kfold_10$lambda_min[results_kfold_10$alpha == best_alpha_min]
lambda_1se <- results_kfold_10$lambda_1se[results_kfold_10$alpha == best_alpha_1se]

k10_lambda_min <- results_k10$lambda_min[a]
k10_lambda_1se <- results_k10$lambda_1se

# Fit final models
enet_model_min <- glmnet(X_train_scaled, y_train, 
                         alpha = 0.5, 
                         lambda = enet_lambda_min)

enet_model_1se <- glmnet(X_train_scaled, y_train, 
                         alpha = 0.5, 
                         lambda = enet_lambda_1se)

# Fitted values and test set predictions
prds.train <- predict(enet_model_min,newx = X_train_scaled, type = "response", s=enet_lambda_min)[,1]
prds.test <- predict(enet_model_min,newx = X_test_scaled, type = "response", s=enet_lambda_min)[,1]

# Prediction Accuracy

library(pROC)
auc.train <- roc(y_train,prds.train)
auc.train

auc.test <- roc(y_test,prds.test)
auc.test

par(mfrow=c(1,2))

par(mfrow=c(1,2))

plot(auc.train)
plot(auc.test)

par(mfrow=c(1,1))




## with k=20 ###

results_k20 <- data.frame()

# Perform grid search, this will test all teh values of alpha then we'll be able to see which one was best
for (alpha_val in alpha_grid) {
  # Cross-validation for current alpha
  cv_model <- cv.glmnet(X_train_scaled, y_train, 
                        alpha = alpha_val, 
                        nfolds = 20,
                        type.measure = "auc", # either auc or deviance, not sure the difference
                        keep = TRUE,
                        family = "binomial")
  
  # Store results
  results_k20 <- rbind(results_k20, data.frame(
    alpha = alpha_val,
    lambda_min = cv_model$lambda.min,
    lambda_1se = cv_model$lambda.1se,
    cvm_min = min(cv_model$cvm),
    cvm_1se = cv_model$cvm[which(cv_model$lambda == cv_model$lambda.1se)],
    nzero_min = cv_model$nzero[which(cv_model$lambda == cv_model$lambda.min)],
    nzero_1se = cv_model$nzero[which(cv_model$lambda == cv_model$lambda.1se)]
  ))
}

# Print results
print(results_k20)

# Find best alpha based on minimum CV error
best_alpha_min <- results_k20$alpha[which.min(results_k20$cvm_min)] # 0.08
best_alpha_1se <- results_k20$alpha[which.min(results_k20$cvm_1se)] # 0.05



## extract top predictors
# Function to extract top N predictors from a model
extract_top_predictors <- function(model, n = 10, lambda_type = "min") {
  # Get coefficients
  if (lambda_type == "min") {
    coefs <- as.matrix(coef(model))[, 1]
  } else {
    # For models stored differently
    coefs <- as.matrix(coef(model, s = paste0("lambda.", lambda_type)))[, 1]
  }
  
  # Remove intercept
  coefs_no_intercept <- coefs[-1]
  
  # Get absolute values for ranking
  abs_coefs <- abs(coefs_no_intercept)
  
  # Get top N predictors
  top_indices <- order(abs_coefs, decreasing = TRUE)[1:min(n, length(abs_coefs))]
  
  # Create result data frame
  result <- data.frame(
    Predictor = names(coefs_no_intercept)[top_indices],
    Coefficient = coefs_no_intercept[top_indices],
    Absolute_Effect = abs_coefs[top_indices],
    Rank = 1:length(top_indices)
  )
  
  return(result)
}

```


To see which are the top predictors:
look at section 5 in regularisation methods comparison.R

```{r}

```








